---
title: "NHS Doctor Income vs Inflation Project"
author: "Hugo Labat"
date: "`r Sys.Date()`"
runtime: shiny
output:
  html_document:
    code_folding: show
    theme: flatly
    highlight: pygments
---
```{r global-options, include=FALSE}
#This sets a few defaults for this entire markdown, to make it visually consistent
#and to prevent messages and warnings without retyping the command.

knitr::opts_chunk$set(echo=TRUE,warning=FALSE,message=FALSE)

```
## Background

The medical profession in the UK revolves around the National Health Service (NHS), a world-renowned system respected for its high-quality and "free at the point of delivery" approach. Part of this system is the training of healthcare professionals, who are mandated to work in the NHS throughout their training in order to be officially recognised in their role. In this lens, however, the NHS becomes a monopsony; where a monopoly is the sole provider of a good or service, a monopsony is the sole buyer.

These above facts, combined with the NHS' chronic lack of funding, has resulted in reduced salaries for all healthcare professionals working in the United Kingdom, when compared to other developed countries. Furthermore, during 2016 junior doctors (FY1 and FY2) held industrial action about the inability for their salaries to match the trends in inflation, as well as the excessive workload they were expected to work under. The income-inflation imbalance lead to effective losses in income and progressively less returns for their almost decade-long investment to education. This protest has had a resurgence, lately, and I was curious to find out why, and if their protests had any effect.


## Aims

I wanted to find out more about the above claims and effects, and aimed to plot the change in NHS doctor salaries versus changes in inflation. To this end, I wanted to create two main line plots to visualise:

1) The cumulative money lost by doctors due to inflation

2) The real salary of doctors versus the salary they should have if it matched changes in inflation

I thought a third, less complicated plot would be helpful in visualising the data, indicating the percentage change in inflation and salary for each month.


## Data Origins

For the salary data, I used the official NHS Earnings Estimates data, which I accessed from [the NHS website (click here for source).](https://web.archive.org/web/20220417162020/https://digital.nhs.uk/data-and-information/publications/statistical/nhs-staff-earnings-estimates) The exact data I used is before and including December 2021, which was the latest version at the time; this included data from September 2009 onwards.

I also took data from June 2007 to December 2008, which was available in pdf format, though this did not have the same staff break-downs as the csv data so I ultimately did not use it, despite extracting the data (as seen later on). The roles I wanted information on are as follows, listed in descending seniority:

- Consultant
- Specialist Registrar (ST)
- Core Trainee (CT)
- Foundation Doctor Year 2 (FY2)
- Foundation Doctor Year 1 (FY1)

This is because these are the mandatory steps for any doctor wanting to specialise in the NHS, with the end-goal being consultancy. Including medical school, this takes a minimum of 10 years.

For the inflation data, I used the Office for National Statistics (ONS) Consumer Price Index data including owner occupiers' Housing costs (CPIH), which I accessed from [the ONS website (click here for source).](https://web.archive.org/web/20220417162131/https://www.ons.gov.uk/economy/inflationandpriceindices/bulletins/consumerpriceinflation/january2022)


## Data-Wrangling

Having downloaded the data, I was ready to extract and process it. All the raw scripts for this are included in my 'docs' folder, and I include the important parts here.


### Setting Up

Below is the set-up needed to run all my code and visualisations. The libraries are stored and taken from renv, to maintain:

- Isolation
- Portability
- Reproducibility

Renv is the R Studio updated version of packrat, and renv.lock can be opened in RStudio to view the versions and packages.

```{r libraries}
#Loading necessary libraries

renv::restore() #To restore packages loaded for this project

library(tidyverse) #To process and graph the data
library(here) #To easily set the working directory
library(pdftools) #To extract the data from the pdf
library(data.table) #To work with the pdf data
library(scales) #To improve visualisation/scales of data
library(shiny) #To create an interactive graph
library(shinyWidgets) #To create nicer interactive graphs
library(ggpubr) #To combine graphs together, with flexible sizes

```


### PDF Data Preparation

Before I knew the pdf data did not contain the break-downs I wanted, I wrote a script to extract the data. This is outlined below.

```{r pdf_prep}
#Creating function to extract salary data from .pdf file
data_extract <- function(raw){
  
  #Splits single pages
  raw <- map(raw,~str_split(.x,"\\n") %>% unlist())
  
  #Concatenate split pages
  raw <- reduce(raw,c)
  
  #Specifying start and end of desired table
  table_start <- stringr::str_which(tolower(raw),"foundation yr 1")-1
  table_end <- stringr::str_which(tolower(raw),"source")-1
  table_end <- table_end[min(which(table_end>table_start))] #this is to make sure I get the right table
  
  #Creating a table & removing the special characters
  table <- raw[(table_start):(table_end)]
  table <- table %>% str_replace_all("\\s{2,}","|") %>%
    str_replace_all("£","") %>% 
    str_replace_all(",","")
  text_con <- textConnection(table)
  data_table <- read.csv(text_con,sep="|")
  
  #Extracting desired column and making numerical
  data_table <- data_table[c(1,4)]
  data_table <- data_table %>% transform(Salary.3=as.numeric(Salary.3))
  
  #Merging consultant role salaries into one row; they occupy rows 4&5
  data_table[c(4,5),2] <- mean(data_table[c(4,5),2])
  data_table <- data.frame(data_table[c(1:4,6:7),c(1,2)])
  
  #Renaming the rows to be code-friendly
  data_table[1] <- c("fy1","fy2","reg","con","spec","staff")
  colnames(data_table) <- c("role","median_salary")
  
  #Adding a new column for the date and rearranging
  data_table$date <- c(NA,NA,NA,NA,NA,NA)
  data_table <- data_table[c(1,3,2)]
  
  #outputting end data frame, so it is use-able by map_df
  data_table
  #end of function
}


#Creating a vector that contains a list of all the files in my pdf folder
pdf_list <- list.files(here("data","raw","pdf"))

#creating df as a master df for all data from all pdfs
pdf_data <- data.frame(role=c("fy1","fy2","reg","con","spec","staff"),
                       date=c(NA,NA,NA,NA,NA,NA),
                       median_salary=c(NA,NA,NA,NA,NA,NA))

#For loop that works through the pdf files in my raw data
for (i in c(1:length(pdf_list))){
  
  #Extracting pdf from folder and turns it into a string of text
  raw_table <- map(here("data","raw","pdf",pdf_list[i]),pdf_text)
  
  #Applying data_extract function on raw data
  data_table <- map_df(raw_table,data_extract)
  
  #adding date to data table
  data_table$date <- sub(".pdf","",pdf_list[i])
  
  #Adding the looped data_table to pdf_data
  pdf_data <- rbind(pdf_data,data_table)
  
  if(i==length(pdf_list)){
    pdf_data <- pdf_data[-c(1:6),]
    rownames(pdf_data) <- c(1:length(pdf_data[,1]))
  }
  #End of loop
}

#Editing date format to be code-friendly, adding 01 to the front to be compatible with as.Date
pdf_data$date <- paste(pdf_data$date,"01",sep="-")
pdf_data$date <- as.Date(pdf_data$date,format="%Y-%m-%d")
pdf_data <- pdf_data[order(pdf_data$date),]

#Saving data to 'refined' data folder
write.csv(pdf_data,here("data","refined","pdf_data.csv"))
```


As is now evident, the data does not contain any information about Core Training years, making it unfit for my project. However, learning how to extract table data from a pdf was invaluable, and I am sure I will use this skill later in my career.


### CSV Data Preparation
Next, I extracted the data from the CSV files. This was comparatively straightforward.

I wanted to extract the salary data titled "basic pay per FTE", which refers to the monthly salary of a doctor per Full Time Equivalent, implying this is what a fully working doctor at that grade would get paid, on average.

```{r csv_salary_prep}
#Extracting data into data frame
salary_df <- read_csv(here("data","raw","csv","2010-08-2021-12.csv"))

#Extracting only date, staff group, payment type, and amount
salary_df <- salary_df %>% select("Date","Staff Group","Payment type","Amount")

#Renaming column titles to be code-friendly
names(salary_df) <- c("date","group","pay_code","amount")


#Creating a vector of the group roles we want to filter in
target_groups <- c("Consultant","Specialty Registrar","Core Training",
                   "Foundation Doctor Year 2","Foundation Doctor Year 1")

#filtering in the desired pay code data
salary_df <- salary_df %>% filter(group %in% target_groups,
                     pay_code=="PUBGRP_010_BASIC_PAY_PER_FTE")

#Removing pay_code since it is now not needed
salary_df <- salary_df[,-c(3)]

#Ensuring 'date' column is considered date data
salary_df$date <- as.Date(salary_df$date,format="%Y-%m-%d")

#creating df to store salary change data from upcoming loop
salary_change_df <- data.frame(NA,NA,NA,NA)
colnames(salary_change_df) <- c("date","group","amount","salary_change")
salary_change_df$date <- as.Date(salary_change_df$date,format="%Y-%m-%d")


#For loop to calculate % change in salary between months for each group
for (i in c(1:length(target_groups))){
  #Creating temporary df organising pay data by group and in date order
  temp_df <- filter(salary_df,group==target_groups[i])
  temp_df <- temp_df[order(temp_df$date),]
  
  #Finding % change in salary per month compared to August 2010, for specified group
  for (val in c(1:nrow(filter(salary_df,group==target_groups[i])))){
    temp_df$salary_change[val] <- as.numeric(100*(temp_df[val,3]-temp_df[1,3])/temp_df[1,3])
  }
  
  #Adding temp_df to salary_change_df
  salary_change_df <- rbind(salary_change_df,temp_df)
  
  #Removing superfluous NA row from salary_change_df
  if (i==length(target_groups)){
    salary_change_df <- salary_change_df[-1,]
    rownames(salary_change_df) <- c(1:length(salary_change_df[,1]))
  }
  #End of loop
}

#Creating df to measure mean salary change across all groups, over time
mean_salary_change_df <- salary_change_df %>% 
  group_by(date) %>%
  summarise(mean_salary_change=mean(salary_change))


#Saving both pieces of data to 'refined' data folder
write.csv(salary_change_df,here("data","refined","salary_change.csv"))
write.csv(mean_salary_change_df,here("data","refined","mean_salary_change.csv"))

```

The CPIH data I extracted is the total CPIH, without the breakdowns:

```{r csv_inflation_prep}
#Extracting data into data frame
cpih_df <- read_csv(here("data","raw","csv","cpih01-time-series-v20.csv"))

#Removing unneeded columns and renaming remaining ones to be code-friendly
cpih_df <- cpih_df[,c(1,2,7)]
names(cpih_df) <- c("cpih","date","code")

#Filtering in the Overall CPIH index data, then removing since it is now not needed
cpih_df <- cpih_df %>% filter(code=="Overall Index")
cpih_df <- cpih_df[,c(2,1)]

#Editing date format to be code-friendly, adding 01 to the front to be compatible with as.Date
cpih_df$date <- paste("01",cpih_df$date,sep="-")
cpih_df$date <- as.Date(cpih_df$date,format="%d-%b-%y")
cpih_df <- cpih_df[order(cpih_df$date),]
cpih_df$cpih <- as.numeric(cpih_df$cpih)

#Removing dates before 2009-09-30, i.e. when the NHS salary data starts
cpih_df <- filter(cpih_df,date >= "2009-09-01")

#Removing cpih dates from 2022-01-01 and 2022-02-01 i.e. when NHS salary data ends
cpih_df <- filter(cpih_df,date <= "2021-12-01")

#Reformatting cpih so that the index is not 2015=100 but 2009-09-01=100
for (i in c(1:nrow(cpih_df))){
  if (i!=nrow(cpih_df)){
    cpih_df$cpih[i+1] <- as.numeric(100+(cpih_df[i+1,2]-cpih_df[1,2]))
  }
  else{
    cpih_df$cpih[1] <- 100
  }
}

#Creating new column for upcoming for loop
cpih_df$cpih_change <- NA

#For loop to calculate the change in cpih between months
for (val in c(1:nrow(cpih_df))){
  if (val==1){
    cpih_df$cpih_change[val] <- 0
  }
  else{
    cpih_df$cpih_change[val] <- as.numeric(cpih_df[val,2]-cpih_df[val-1,2])
  }
}


#Saving data to 'refined' data folder
write.csv(cpih_df,here("data","refined","cpih_data.csv"))

```

With both sets of data extracted and made code-friendly, I was ready to combine them and create graphs:

```{r income_vs_inflation}
#Reformatting cpih so that the index starts at 0, to match salary change dfs
cpih_df$cpih <- cpih_df$cpih-100

#Creating a cpih df to be compatible with all groups of salary_change_df
for (i in c(1:5)){
  if (i==1){
    cpih_all_groups <- cpih_df
  }
  else{
    cpih_all_groups <- rbind(cpih_all_groups,cpih_df)
  }
}


#First Data-Wrangling:
#Comparing mean salary data with cpih data

#Merging mean data and cpih data & removing superfluous date column
cpih_vs_mean_salary <- cbind(mean_salary_change_df,cpih_df)
cpih_vs_mean_salary <- cpih_vs_mean_salary[,-3]

#Calculating difference between cpih and mean salary change
cpih_vs_mean_salary$diff <- cpih_vs_mean_salary$mean_salary_change-cpih_vs_mean_salary$cpih



#Second Data-Wrangling:
#Calculating effective money gained/lost from beginning to end of time sequence
#For average income of all doctors

#The comparison income will be the mean income per month of all doctors
mean_income <- mean(salary_change_df$amount)

#Altering cpih_vs_mean_salary to include a column with cumulative income change
for (i in c(1:nrow(cpih_vs_mean_salary))){
  if (i==1){
    cpih_vs_mean_salary$cumulative_income_change[1] <- 0 #set baseline
  }
  else{
    cpih_vs_mean_salary$cumulative_income_change[i] <- 
      (mean_income*(cpih_vs_mean_salary$diff[i]/100))+
      cpih_vs_mean_salary$cumulative_income_change[i-1]
  }
}

#Changing data to represent cumulative income lost due to inflation
cpih_vs_mean_salary$cumulative_income_lost <- -cpih_vs_mean_salary$cumulative_income_change



#Third Data-Wrangling:
#Comparing per-group salary data with cpih data

#Merging all groups data and cpih data & removing superfluous date column
cpih_vs_salary <- cbind(salary_change_df,cpih_all_groups)
cpih_vs_salary <- cpih_vs_salary[,-5]

#Calculating difference between cpih and salary change
cpih_vs_salary$diff <- cpih_vs_salary$salary_change-cpih_vs_salary$cpih



#Calculating the number of rows in one group, so as to avoid hard-coding values
#Since the number of rows is the same per group, I need only measure the first group
group_size <- nrow(cpih_vs_salary %>% filter(group==cpih_vs_salary$group[1]))



#Fourth Data-Wrangling:
#Calculating effective money gained/lost from beginning to end of time sequence
#For each income of each doctor group
for (i in c(1:nrow(cpih_vs_salary))){
  if ((i-1)%%group_size==0){
    cpih_vs_salary$cumulative_income_lost[i] <- 0 #set baseline for each group
  }
  else{
    for (val in c(1:5)){
      if ((i-1)<=(val*group_size) & (i-1)%/%group_size==(val-1)){
        cpih_vs_salary$cumulative_income_lost[i] <- 
          -((cpih_vs_salary$amount[(group_size*(val-1))+1]*(cpih_vs_salary$diff[i]/100))-
              cpih_vs_salary$cumulative_income_lost[i-1])
      }
    }
  }
}

#Ordering the 'group' column in order of descending seniority (consultant->fy1)
cpih_vs_salary$group <- factor(cpih_vs_salary$group,
                               levels=c("Consultant",
                                        "Specialty Registrar",
                                        "Core Training",
                                        "Foundation Doctor Year 2",
                                        "Foundation Doctor Year 1"))



#FIfth Data-Wrangling:
#Calculating how much doctors should be earning if their incomes consistently
#increased with cpih since 2009, and comparing this to present income changes

#Using cpih_change from cpih_df to calculate cpih-adjusted salary
for (i in c(1:nrow(cpih_vs_salary))){
  if (i==1){
    cpih_vs_salary$cpih_matched_salary[i] <- cpih_vs_salary$amount[i] #set beginning baseline
  }
  else if ((i-1)%%group_size==0){
    cpih_vs_salary$cpih_matched_salary[i] <- cpih_vs_salary$amount[i] #set baseline for each group
  }
  else{
    cpih_vs_salary$cpih_matched_salary[i] <-
      (cpih_vs_salary$cpih_matched_salary[i-1]*(cpih_vs_salary$cpih_change[i])/100)+
      cpih_vs_salary$cpih_matched_salary[i-1]
  }
}


#Saving data to 'refined' data folder
write.csv(cpih_vs_salary,here("data","refined","cpih_vs_salary.csv"))

```

Each step of data-wrangling is important for the overall visualisation of the data, though I eventually opted to use only the last three steps for my graphs. This is because mean data neglects the large differences between groups, and my aims are to compare groups. In the **"docs"** folder, my scripts produce graphs for each step of data-wrangling, though my final visualisation is below.


## Visualisation

I opted for an interactive R Shiny application, as I believe it gives a better representation of the change over time (due to the option to animate), and lets the user tangibly interact with the data.

I aimed to make the visualisation accessible to people with colour-blindness, and as visually simple but elegant as possible. I tried to include subtle visual indications that would help the user interact with the graph, such as thicker lines for main breaks and a neutral grey background.

As previously mentioned, I included a bonus, smaller graph to indicate the percentage change in salary and inflation, as I feel it gives the user a better intuitive grasp of the data; it helped me understand the reasons behind the trends seen in the line graphs.

```{r shiny_app,results="hide"}
#Creating df with columns for change, date, and groups only, adding cpih to 'group' column
col_df <- data.frame(cpih_vs_salary$date,cpih_vs_salary$group,cpih_vs_salary$salary_change)
colnames(col_df) <- c("date","group","change")

temp_cpih_df <- data.frame(cpih_vs_salary$date,cpih_vs_salary$cpih)
temp_cpih_df <- temp_cpih_df[c(1:group_size),]
temp_cpih_df$group <- "CPIH"
temp_cpih_df <- temp_cpih_df[,c(1,3,2)]
colnames(temp_cpih_df) <- c("date","group","change")

col_df <- rbind(col_df,temp_cpih_df)



#Creating R Shiny interactive elements:

#Selector of dates to display
sliderTextInput(inputId="date",
                label="Select Date:",
                choices=as.Date(col_df$date[c(1:group_size)]), #list of months
                selected=as.Date(col_df$date[group_size]),
                grid=TRUE,
                animate=animationOptions(interval=800),
                width="100%"
                )

#Selector of graphs to display
selectInput(inputId="plot",
            label="Select Desired Plot:",
            choices=c("Cumulative Salary Lost to Inflation",
                      "Real vs Inflation-Matched Salary"),
            width="100%"
            )

#Reactive expression to speed up update of plotted data and thus of shiny graph
filtered_df <- reactive({
    cpih_vs_salary %>% filter(date<=input$date)
})


#Plotting both cpih_vs_salary and col_df onto output$plot using ggarrange
#Using if statements to allow selectInput to switch between main plots
  
observeEvent(input$plot,{
  if(input$plot=="Cumulative Salary Lost to Inflation"){
    output$plot <- renderPlot({
    
    #Plotting cumulative salary loss, using filtered_df() & geom_line
    loss_plot <- ggplot()+
      geom_line(data=filtered_df(),
                aes(x=date,
                    y=cumulative_income_lost,
                    colour=group),
                size=1)+
      theme_light()+
      scale_y_continuous(labels=dollar_format(prefix="£"),
                         breaks=seq(0,90000,10000),
                         minor_breaks=seq(-4000,100000,2000),
                         limits=c(-4000,94000),
                         expand=c(0,0))+
      scale_x_date(date_breaks="1 year",
                   date_minor_breaks="3 months",
                   date_labels="%Y",
                   limits=as.Date(c("2009-07-01","2022-06-15")),
                   expand=c(0,0))+
      labs(title="Cumulative Salary Lost to Inflation",
           subtitle=paste("Range: Sep 2009 -",format(as.Date(input$date),"%b %Y"),
                          "; Sep 2009 = 0"))+
      theme(plot.title=element_text(colour="#56B4E9",face="bold"),
            panel.grid.major=element_line(size=1,colour="grey85"),
            panel.grid.minor=element_line(colour="grey87"),
            axis.title.x = element_text(vjust=-0.35),
            axis.title.y = element_text(vjust=2.5),
            legend.position="none", #remove legend, since col_plot has one already
            axis.ticks=element_line(size=1))+ #to match major grid size
      xlab("Date")+
      ylab("Cumulative Salary Loss")+
      scale_colour_manual(values=c("#E69F00",
                                   "#56B4E9",
                                   "#009E73",
                                   "#F0E442",
                                   "#D55E00")) #compatible for colour-blind people
    
    #Plotting monthly % changes in salary and inflation, using col_df & geom_col
    col_plot <- ggplot()+
      geom_col(data=col_df %>% filter(date==input$date),
               aes(x=group,y=change,fill=group))+
      scale_fill_manual(values=c("#E69F00",
                                 "#56B4E9",
                                 "#009E73",
                                 "#F0E442",
                                 "#D55E00",
                                 "#888888"),
                        labels=c("Consultant (CCT)",
                                 "Specialist Trainee (ST)",
                                 "Core Trainee (CT)",
                                 "Foundation Doctor Year 2 (FY2)",
                                 "Foundation Doctor Year 1 (FY1)",
                                 "Consumer Price Index with Housing (CPIH)"))+
      theme_light()+
      labs(title="Change in Salary and Inflation",
           subtitle=paste("Date:",format(as.Date(input$date),"%b %Y"),
                          "; Sep 2009 = 0"),
           fill="Group")+
      theme(plot.title=element_text(colour="#56B4E9",face="bold"),
            panel.grid.major=element_line(size=1,colour="grey85"),
            panel.grid.minor=element_line(colour="grey87"),
            axis.ticks=element_line(size=1),
            axis.title.x = element_text(vjust=-0.35),
            axis.title.y = element_text(vjust=2.5))+
      xlab("Group")+
      ylab("Percentage Change")+
      scale_x_discrete(labels=c("CCT","ST","CT","FY2","FY1","CPIH"))+
      scale_y_continuous(limits=c(-5,38),
                         breaks=seq(0,35,5),
                         minor_breaks=seq(-8,38,1),
                         expand=c(0,0))
    
    #Using ggarrange to merge the plots into one
    ggarrange(loss_plot,
              col_plot,
              ncol=2,nrow=1)
    })
  }
  
  else{
    output$plot <- renderPlot({
      #Plotting monthly % changes in salary and inflation, using col_df & geom_col
      col_plot <- ggplot()+
        geom_col(data=col_df %>% filter(date==input$date),
        aes(x=group,y=change,fill=group))+
        scale_fill_manual(values=c("#E69F00",
                                   "#56B4E9",
                                   "#009E73",
                                   "#F0E442",
                                   "#D55E00",
                                   "#888888"),
                          labels=c("Consultant (CCT)",
                                   "Specialist Trainee (ST)",
                                   "Core Trainee (CT)",
                                   "Foundation Doctor Year 2 (FY2)",
                                   "Foundation Doctor Year 1 (FY1)",
                                   "Consumer Price Index with Housing (CPIH)"))+
        theme_light()+
        labs(title="Change in Salary and CPIH",
             subtitle=paste("Date:",format(as.Date(input$date),"%b %Y"),
                            "; Sep 2009 = 0"),
             fill="Group")+
        theme(plot.title=element_text(colour="#56B4E9",face="bold"),
              panel.grid.major=element_line(size=1,colour="grey85"),
              panel.grid.minor=element_line(colour="grey90"),
              axis.ticks=element_line(size=1),
              axis.title.x = element_text(vjust=-0.35),
              axis.title.y = element_text(vjust=2.5))+
        xlab("Group")+
        ylab("Percentage Change")+
        scale_x_discrete(labels=c("CCT","ST","CT","FY2","FY1","CPIH"))+
        scale_y_continuous(limits=c(-5,38),
                           breaks=seq(0,35,5),
                           minor_breaks=seq(-8,38,1),
                           expand=c(0,0))
        
        
        #Plotting real vs inflation-matched salary, using filtered_df() & geom_line
        match_plot <- ggplot()+
          geom_line(data=filtered_df(),
                    aes(x=date,
                        y=amount,
                        colour=group),
                    size=1)+
          geom_line(data=filtered_df(),
                    aes(x=date,
                        y=cpih_matched_salary,
                        colour=group),
                    size=1,
                    linetype=2)+
          theme_light()+
          scale_y_continuous(labels=dollar_format(prefix="£"),
                             breaks=seq(2000,9000,1000),
                             minor_breaks=seq(1000,10000,200),
                             limits=c(1400,9800),
                             expand=c(0,0))+
          scale_x_date(date_breaks="1 year",
                       date_minor_breaks="3 months",
                       date_labels="%Y",
                       limits=as.Date(c("2009-07-01","2022-06-15")),
                       expand=c(0,0))+
          labs(title="Real vs Inflation-Matched Salary",
               subtitle=paste("Range: Sep 2009 -",format(as.Date(input$date),"%b %Y"),
                              "; - - - = Inflation-Matched, ── = Real Salary"),
               colour="Stage of Training")+
          theme(plot.title=element_text(colour="#56B4E9",face="bold"),
                panel.grid.major=element_line(size=1,colour="grey85"),
                panel.grid.minor=element_line(colour="grey90"),
                axis.title.x = element_text(vjust=-0.35),
                legend.position="none", #remove legend, since col_plot has one already
                axis.title.y = element_text(vjust=2.5), #shifting axis titles to give space
                axis.ticks=element_line(size=1))+ #to match major grid size
          xlab("Date")+
          ylab("Monthly Salary")+
          scale_colour_manual(values=c("#E69F00",
                                       "#56B4E9",
                                       "#009E73",
                                       "#F0E442",
                                       "#D55E00")) #compatible for colour-blind people
        
        #Using ggarrange to merge the plots into one
        ggarrange(match_plot,
                  col_plot,
                  ncol=2,nrow=1)
    })
    }
  })

#Main output
plotOutput(outputId="plot")

```

The above code provides the following visualisation of data from September 2009 to December 2021:

```{r interactive_graph,echo=FALSE}
#Selector of dates to display
sliderTextInput(inputId="date",
                label="Select Date:",
                choices=as.Date(col_df$date[c(1:group_size)]), #list of months
                selected=as.Date(col_df$date[group_size]),
                grid=TRUE,
                animate=animationOptions(interval=800),
                width="100%",
                )

#Selector of graphs to display
selectInput(inputId="plot",
            label="Select Desired Line Graph:",
            choices=c("Cumulative Salary Lost to Inflation",
                      "Real vs Inflation-Matched Salary"),
            width="100%"
            )

#Main output
plotOutput(outputId="plot",
           width="100%")

```

**The above interactive plot is the main visualisation of my project**

All elements are made to be viewed and interacted with via a wide screen. Whilst they would certainly work on a mobile device, much of the data would be compressed and less aesthetically pleasing.

This was a difficult but rewarding experience, and I truly enjoyed learning how to code the app and upload it onto the shinyapps.io website.


## Interpretation

In the *"Cumulative Salary Lost to Inflation"* graph, any gradient above zero indicates an increase in money lost to inflation. The graph provides further intuition by showing the overall trend, and it can be seen that only two groups (FY1 and CT) are getting paid better than inflation. The worst affected group by far are the Consultants.

In the *"Real vs Inflation-Matched Salary"* graph, it can be seen how the CT group is the only group with income that substantially improves compared to inflation, from mid 2019 onwards. FY1s are being paid enough to match inflation, and have been since 2017. Every other group has a marked difference in what they are getting paid versus what they should be paid if pay matched inflation.


## Discussion

Looking at the *"Real vs Inflation-Matched Salary"* graph, the first thing I noticed was the dips in salary every August. In discussing with qualified physicians, I learned that this is when doctors start their new posts; evidently the NHS reduces the salary due to the fact that these doctors start on the first Wednesday and not the first day of the month, though this corrects itself by September.

The second point I noticed in both graps was how the 'entrance roles' (i.e. FY1 and CT) have incomes that now match - and even exceed - inflation. However, the roles one becomes locked into (i.e. FY2, ST, and Consultancy) continue to receive changes in income that fall below the changes in inflation. It is hard to see why this would be done accidentally; this could be the NHS' way of keeping costs down whilst - on the surface - appearing to match inflation with their salaries; by the time doctors encounter the non-matched salaries, they are already locked into their training roles, and leaving would have dire consequences on their career.

On the more pleasant side, the 2016 lobbying and strikes by Junior Doctors (FY1 & FY2) seems to have partially worked in improving their pay, as evidenced by the reversal of the *"Cumulative Salary Lost to Inflation"* line for FY1 doctors.

Finally, Consultants are the group most negatively affected by inflation; if a consultant received mean pay from September 2009 to December 2021, they will have lost £91,257 to inflation. This makes up over a year's salary for most consultants.


## Possible Extensions

From this visualisation, I envision four additional projects that could complement it:

**1) Creating a function that automatically extracts the desired data (specified above), adjusts it according to the function inputs, and outputs a Shiny App.** This is already almost done, and requires only a few changes to the above scripts to achieve (i.e. flexibility of data inputs and graph limits). This function could be applied to a script that automatically downloads the relevant data files, providing a live visualisation of the latest inflation and salary data.

**2) Comparing the salary vs inflation data of the UK to other countries around the world.** This may require one to use the mean data extracted in the first and second data-wrangling sections, and a few adjustments to ensure the data being compared is equal (e.g., accounting for different working hours).

**3) Delving into why the NHS caused the three large peaks and troughs in salary from 2019 to 2021.** Whilst this trend was noticed for most August dates (especially in the *Real vs Inflation-Matched graph*), it would be interesting to see a more detailed breakdown of salaries to see what part of their salary dropped (i.e., was it fewer bonuses because the new staff don't work until a few days in?).

**4) Investigating other government and private bodies to see if they also show a marked drop in salary during the influx of new employees.** Such a trend would be useful to those potential new employees, and would be the grounds for further investigation into how payment differs depending on if you are a new employee or not (i.e., do you get the drop, even if you aren't a new employee?).


## GitHub Repository

You can find the repository for this project by [clicking here](https://github.com/hvlabat/nhs_income_vs_inflation).